---
title: "STA 610L: Module 1.1"
subtitle: "Course overview"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome!



---

## What is this course about?

<i class="fa fa-book fa-2x"></i> Learning the foundations of hierarchical modeling.

--

<i class="fa fa-folder-open fa-2x"></i> Working through the theory of some hierarchical models.

--

<i class="fa fa-tasks fa-2x"></i> Using hierarchical models to answer inferential questions.

--

<i class="fa fa-database fa-2x"></i> Applying hierarchical models to real datasets.

--

<i class="fa fa-group fa-2x"></i> Honing collaborative and presentations skills.

--

--- 
<i class="fa fa-quote-left fa-2x fa-pull-left fa-border" aria-hidden="true"></i>
<i class="fa fa-quote-right fa-2x fa-pull-right fa-border" aria-hidden="true"></i>
The Bayesian paradigm is well suited for building hierarchical models. Usually you just have several levels of conditional distributions making up the prior.



---
## Instructor

[Dr. Olanrewaju Michael Akande](https://olanrewajuakande.com)

<i class="fa fa-envelope"></i> &nbsp; [olanrewaju.akande@duke.edu](mailto:olanrewaju.akande@duke.edu) <br>
<i class="fa fa-home"></i> &nbsp; [https://sta610-f21.olanrewajuakande.com](https://sta610-f21.olanrewajuakande.com) <br>
<i class="fa fa-calendar"></i> &nbsp; Mondays and Thursdays (9am -- 10am) <br>
<i class="fa fa-university"></i> &nbsp; Zoom Meeting ID: **See Sakai**


---
## TAs

[Chengxin Yang](https://scholars.duke.edu/person/chengxin.yang)

<i class="fa fa-envelope"></i> &nbsp; [chengxin.yang@duke.edu](mailto:chengxin.yang@duke.edu) <br>
<i class="fa fa-calendar"></i> &nbsp; Tuesdays (1pm - 2pm) and Fridays (4pm - 5pm) <br>
<i class="fa fa-university"></i> &nbsp; Zoom Meeting ID: **See Sakai**

[Alessandro Zito](https://scholars.duke.edu/person/alessandro.zito)

<i class="fa fa-envelope"></i> &nbsp; [alessandro.zito@duke.edu](mailto:alessandro.zito@duke.edu) <br>
<i class="fa fa-calendar"></i> &nbsp; Wednesdays (10am - 11am) and Fridays (8:30am - 9:30am) <br>
<i class="fa fa-university"></i> &nbsp; Zoom Meeting ID: **See Sakai**



---
## FAQs

All materials and information will be posted on the course website.

--

- How much theory will this class cover? *A decent amount.*

--

- Am I prepared to take this course? *Yes, if you are familiar with the topics covered in STA 360/601/602 (Bayesian Inference) and all its prerequisite at Duke.*

--

- What if I can't remember the topics in the prerequisites? *See the review materials in the next module.*

--

- Will we be doing "very heavy" computing? *Not too heavy but a good amount.*

--

- What computing language will we use? *R!*

--

- What if I don't know R? *This course assumes you already know R but you can still learn on the fly (you must be self-motivated). Here are some resources for you: [https://sta610-f21.olanrewajuakande.com/resources/](https://sta610-f21.olanrewajuakande.com/resources/).*




---

class: center, middle

# Course structure and policies


---
## Course structure and policies

- See: [https://sta610-f21.olanrewajuakande.com/policies/](https://sta610-f21.olanrewajuakande.com/policies/)

--

- Make use of the teaching team's office hours, we're here to help!

--

- Do not hesitate to show up during office hours. You can also make an appointment to discuss a homework problem or any aspect of the course. 

--

- When the teaching team has announcements for you we will send an email to your Duke email address. Please make sure to check your email daily.

--

- Try as much as possible to refrain from texting or using your computer for anything other than coursework while in class.


---
## Other details

- What topics will we cover? Refer to Section 11 of the syllabus (here: [Syllabus](https://sta610-f21.olanrewajuakande.com/syllabus_pdf/Syllabus.pdf)).

--

- Also, refer to the schedule on the website for an updated breakdown of the topics. Remember to refresh the page frequently. See here: [Class Schedule](https://sta610-f21.olanrewajuakande.com/syllabus/).

--

- If you are auditing this course, remember to complete the necessary audit forms.

--

- Confirm that you have access to Sakai, Ed Discussion and Gradescope.

--

- Finally, please respect Duke's masking and social distancing policies, both in and out of classes.



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




