---
title: "STA 610L: Module 3.2"
subtitle: "Linear mixed effects models (influence measures)"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
#library(tufte)
#library(sjPlot)
library(lme4)
#library(sjstats)
#library(merTools)
#library(lattice)
library(ggplot2)
library(lattice)
library(influence.ME)
```



## Residuals

Residual analysis and diagnostic methods are well developed for linear regression models (c.f., Cook and Weisberg, 1982), but they are somewhat less developed for mixed effects models.

<br>

The following slides are based on [Nieuwenhuis et al](https://journal.r-project.org/archive/2012-2/RJournal_2012-2_Nieuwenhuis~et~al.pdf).



---
## Example: Orthodontics Data

Consider the dental data with model

$$Y_{ij}=\beta_0+\beta_1I(\text{male})_i+\beta_2t_j+\beta_3I(\text{male})_it_j + b_{0i} + b_{1i}t_j + \varepsilon_{ij},$$ 

where
$$\begin{pmatrix} b_{0i} \\ b_{1i} \end{pmatrix} \overset{iid}\sim N\left(0,\begin{pmatrix}d_{11} & d_{12} \\ d_{12} & d_{22}\end{pmatrix}\right) \perp \varepsilon_{ij} \overset{iid}\sim N(0,\sigma^2),$$
for illustration.



---
## Residuals

Generally, the .hlight[residuals] $y_{ij}-\widehat{y}_{ij}$ can be helpful in flagging outliers and in assessing the adequacy of most fitted models.

However, the definition of residuals is not always consistent in the case of mixed effects or hierarchical models:

  - Many texts define residuals for subject/group $i$ as $Y_i-X_i\widehat{\beta}$. 
  
  - Many software implementations define residuals as $Y_i-X_i\widehat{\beta}-Z_i\widehat{b}_i$ (nice because these can then be analyzed using standard methods)
  
These are easy to compute and we already did the later in a previous module.


---
## Review: residual analysis

That said, in any case, residual analysis is not always a great tool for detecting .hlight[influential] cases:

  - Cases with high residuals or high standardized residuals are called .hlight[outliers]
  
  - Outliers may or may not be influential in the model fit
  
  - An influential case may dominate the regression model so that the line is drawn more closely towards the case (making it an .hlight[inlier])



---
## Review: influence

We hope that all data points have some amount of influence on our parameter estimates. 

However, we may be concerned if a single case has disproportionate influence on model results. 

If so, one observation or group of observations may pull the estimated regression line towards the group. 

In such a case, excluding a single group might have a substantial effect on estimates. 

This idea is behind the development of many popular influence diagnostics, often termed .hlight[deletion diagnostics].





---
## Review: leverage

The degree to which an observation has the *potential* to be influential is closely related to the .hlight[leverage] of the case, which is a measure of how extreme the case is in the $X$ space.

Leverage is not simply defined as an outlying value in $X$ space of a single variable but also in a multivariate sense.

For example, in a study of pregnancy outcomes, it may be relatively common to have mothers who are 40, or fathers who are 20, but babies who have a 40 year old mother and a 20 year old father may be fairly uncommon.

As you should already know, the leverage score for an observation $i$ is the $i$th diagonal element of the projection or hat matrix.



---
## Influence

It is not necessarily the case that outliers or cases with high leverage are influential.

So, how do we detect influential cases?

One popular approach is to use the principle that when a single case is removed from the data entirely, we would like for models based on the data not to give vastly different conclusions.

If parameter estimates change a lot after a single individual is excluded, then the individual may be considered .hlight[influential].



---
## Multi-level influence

Mixed effects and multilevel models estimate effects of lower-level and higher-level variables.

It is thus possible that in some cases a higher-level group is influential (more likely when you don't have very many groups), while in others, a single observation within a group is influential.

We will examine influence at both levels.



---
## Influence of higher-Level observations

.hlight[DFBETAS] (standardized difference of the beta) measures the level of influence observations have on single parameter estimates.

It is calculated as the difference in magnitude of the parameter estimate between the model including and the model excluding the group (or individual in a longitudinal study), standardized by dividing by the standard error of the estimate that excludes the group (to prevent variance inflation from masking the level of influence).  

For group $i$ and parameter $k$,
$$\text{DFBETAS}_{ik}=\frac{\widehat{\gamma}_k-\widehat{\gamma}_{k(-i)}}{se(\widehat{\gamma}_{k(-i)})},$$

where $\widehat{\gamma}_k$ is the original estimate of the $k$th parameter, and $\widehat{\gamma}_{k(-i)}$ is the estimate of the same parameter after group $i$ has been excluded from the data.



---
## DFBETAS

Belsley (1980) recommends a cutoff of $\frac{2}{\sqrt{n}}$ for identifying overly influential observations.

Here $n$ is defined as the number of groups at the level of removal $(-i)$ for the calculation.

(For the dental data we have 27 kids and 4 observations per kid, so at the group level $n=27$.)

```{r echo=FALSE}
data(Orthodont,package="nlme")
Orthodont$Sex=relevel(Orthodont$Sex,ref="Female")
m1 <- lmer(distance~Sex+age+age*Sex+(1+age|Subject),data=Orthodont)
```


```{r eval=FALSE}
library(influence.ME)
m1.inf <- influence(m1,"Subject")
#use obs argument for observation-level deletion
print(2/sqrt(length(unique(Orthodont$Subject))))
dfbetas(m1.inf)
#note that there can be issues with singularity when we start removing groups
```



---
## DFBETAS

.small[
```{r dfbetas2,warning=FALSE,message=FALSE}
m1.inf <- influence(m1,"Subject"); print(2/sqrt(length(unique(Orthodont$Subject))))
round(dfbetas(m1.inf),4)
```
]

Here we see that M04 and M13 are somewhat influential on some of our estimates. What did these kids look like?



---
## DFBETAS

```{r plotdfbetas, fig.height=4.7}
plot(m1.inf,which="dfbetas",xlab="DFBETAS",ylab="Student")
```



---
## DFBETAS

```{r m0413,fig.height=4.5,message=FALSE,warning=FALSE}
plot(m1, distance ~ fitted(.) | Subject, abline = c(0,1))
```




---
## DFBETAS

```{r reminder}
Orthodont$distance[Orthodont$Subject=="M04"]
Orthodont$distance[Orthodont$Subject=="M13"]
```

M04 had large measurements without a lot of growth over time -- pulling him out of the model reduced the intercept for boys and also decreased their slope.
  
M13 had a small measure at age 8 and then grew substantially. Leaving him out of the model changed the estimates significantly, greatly increasing the intercept for boys and also reducing the slope among boys.
  



---
## Cook's distance

When the number of observations or predictors is large, it may take a while to get through all the DFBETAS.

.hlight[Cook's distance] gives us a summary measure for influence on all parameter estimates.

It is defined as
$$C_i=\frac{1}{p}(\widehat{\gamma}-\widehat{\gamma}_{(-i)})'\widehat{\Sigma}_{(-i)}^{-1}(\widehat{\gamma}-\widehat{\gamma}_{(-i)})$$
where $p$ is the length of $\beta$, and $\widehat{\Sigma}_{(-i)}$ is the covariance matrix of the parameter estimates excluding group $i$.

Van der Meer et al (2010) recommends a cutoff of $\frac{4}{n}$ where again $n$ is the number of groups in the grouping factor being evaluated.

If there is just one parameter in the model, then Cook's distance is the DFBETAS squared for that parameter.



---
## Cook's distance

.large[
```{r cd}
print(4/length(unique(Orthodont$Subject)))
cooks.distance(m1.inf,sort=TRUE)
```
]


---
## Cook's distance

```{r cd3, fig.height=4}
plot(m1.inf,which="cook",cutoff=4/length(unique(Orthodont$Subject)), 
     sort=TRUE,xlab="Cook's D",ylab="Subject")
```

It's M13 again.



---
## Other metrics

There are many other metrics we could use.

One option is the .hlight[percentile change], which is defined as
$$\frac{\widehat{\gamma}-\widehat{\gamma}_{(-i)}}{\widehat{\gamma}}\times 100%$$



---
## Other metrics

```{r pchange, fig.height=4.5}
plot(m1.inf,which="pchange",xlab="percentile change",ylab="Student")

```




---
## Other metrics

Another metric is the .hlight[changes in significance].

Basically, we evaluate whether excluding a group changes the statistical significance of any of the estimates in the model.

The user sets the critical value, and estimates that did not exceed it but do so when the group is removed, or *vice versa*, are flagged.

See the .hlight[sigtest] function.


---
## Influence of lower-Level observations

We can also look at the influence of single lower-level observations.

They could be impactful in longitudinal data for example, when we have relatively few observations per individual.

Note however that the computational complexity of these deletion diagnostics will be increased in this case.

Here we look at Cook's Distance for the dental data on the individual observation level:

```{r cookindiv, message=FALSE,warning=FALSE}
m1.inf.indiv <- influence(m1,obs=TRUE)
m1.cook <- cooks.distance(m1.inf.indiv)
infindiv <- m1.cook>4/length(Orthodont$distance)
```



---
## Influence of lower-Level observations

.small[
```{r cookindiv2,message=FALSE,warning=FALSE}
data.frame(Orthodont$Subject,m1.cook,infindiv)[1:35,]
```
]


---
## Influence of lower-Level observations

.small[
```{r cookindiv3,message=FALSE,warning=FALSE}
data.frame(Orthodont$Subject,m1.cook,infindiv)[36:72,]
```
]



---
## Influence of lower-Level observations

.small[
```{r cookindiv4,message=FALSE,warning=FALSE}
data.frame(Orthodont$Subject,m1.cook,infindiv)[73:108,]
```
]

M13 once again!




---
## Dealing with influential data

What to do with influential data is a much harder problem.

Reasonable strategies may include the following.

  - Verify data recorded correctly
  
  - Consider robust models
  
  - Determine whether any lurking predictors should be added to the model
  
  - Report results with and without overly influential results
  


---

class: center, middle

# What's next?

### Move on to the readings for the next module!




