---
title: "STA 610L: Module 3.6"
subtitle: "Logistic mixed effects model (Part II)"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(rvest)
#library(tufte)
#library(sjPlot)
library(lme4)
#library(sjstats)
#library(merTools)
#library(lattice)
#library(ggplot2)
#library(brms)
#library(dutchmasters)
#library(tidyverse)
#library(rethinking)
library(arm)
```



## 1988 elections analysis

The dataset includes 2193 observations from one of eight surveys (the most recent CBS News survey right before the election) in the original full data.

.small[
Variable    | Description
:------------- | :------------
org | cbsnyt = CBS/NYT
.hlight[bush] | .hlight[1 = preference for Bush Sr., 0 = otherwise]
state | 1-51: 50 states including DC (number 9)
edu | education: 1=No HS, 2=HS, 3=Some College, 4=College Grad
age | 1=18-29, 2=30-44, 3=45-64, 4=65+
female | 1=female, 0=male
black | 1=black, 0=otherwise
region | 1=NE, 2=S, 3=N, 4=W, 5=DC
v_prev | average Republican vote share in the three previous elections (adjusted for home-state and home-region effects in the previous elections)
]

Given that the data has a natural multilevel structure (through `state` and `region`), it makes sense to explore hierarchical models for this data.



---
## 1988 elections analysis

Both voting turnout and preferences often depend on a complex combination of demographic factors.

In our example dataset, we have demographic factors such as biological sex, race, age, education, which we may all want to look at by state, resulting in $2 \times 2 \times 4 \times 4 \times 51 = 3264$ potential categories of respondents.

We may even want to control for `region`, adding to the number of categories.

Clearly, without a very large survey (most political survey poll around 1000 people), we will need to make assumptions in order to even obtain estimates in each category.

We usually cannot include all interactions; we should therefore select those to explore (through EDA and background knowledge).

The data is in the file `polls_subset.txt` on Sakai.



---
## 1988 elections analysis

```{r}
###### Load the data
polls_subset <- read.table("data/polls_subset.txt",header=TRUE)
str(polls_subset)
head(polls_subset)
```


---
## 1988 elections analysis

```{r}
summary(polls_subset)
```



---
## 1988 elections analysis

```{r}
polls_subset$v_prev <- polls_subset$v_prev*100 #rescale 
polls_subset$region_label <- factor(polls_subset$region,levels=1:5,
                                    labels=c("NE","S","N","W","DC"))
#we consider DC as a separate region due to its distinctive voting patterns
polls_subset$edu_label <- factor(polls_subset$edu,levels=1:4,
                                 labels=c("No HS","HS","Some College","College Grad"))
polls_subset$age_label <- factor(polls_subset$age,levels=1:4,
                                 labels=c("18-29","30-44","45-64","65+"))
#the data includes states but without the names, which we will need, 
#so let's grab that from R datasets
data(state) 
#"state" is an R data file (type ?state from the R command window for info)
state.abb #does not include DC, so we will create ours
#In the polls data, DC is the 9th "state" in alphabetical order
state_abbr <- c (state.abb[1:8], "DC", state.abb[9:50])
polls_subset$state_label <- factor(polls_subset$state,levels=1:51,labels=state_abbr)
rm(list = ls(pattern = "state")) #remove unnecessary values in the environment
```



---
## 1988 elections analysis

```{r}
###### View properties of the data  
head(polls_subset)
dim(polls_subset)
```



---
## 1988 elections analysis

```{r}
###### View properties of the data  
str(polls_subset)
```




---
## 1988 elections analysis

I will not do any meaningful EDA here.

I expect you to be able to do this yourself.

Let's just take a look at the amount of data we have for "bush" and the age:edu interaction.

```{r}
###### Exploratory data analysis
table(polls_subset$bush) #well split by the two values
table(polls_subset$edu,polls_subset$age)
```


---
## 1988 elections analysis

As a start, we will consider a simple model with fixed effects of race and sex and a random effect for state (50 states + the District of Columbia).
$$
\begin{split}
\textrm{bush}_{ij} | \boldsymbol{x}_{ij} & \sim \textrm{Bernoulli}(\pi_{ij}); \ \ \ i = 1, \ldots, n; \ \ \ j = 1, \ldots, J=51; \\
\textrm{log}\left(\dfrac{\pi_{ij}}{1-\pi_{ij}}\right) & = \beta_{0} + b_{0j} + \beta_1 \textrm{female}_{ij} + \beta_2 \textrm{black}_{ij}; \\
b_{0j} & \sim N(0, \sigma^2).
\end{split}
$$

In `R`, we have
```{r results = 'hide'}
#library(lme4)
model1 <- glmer(bush ~ black+female+(1|state_label),
                family=binomial(link="logit"),
                data=polls_subset)
summary(model1)
```



---
## 1988 elections analysis
```{r echo=F}
summary(model1)
```


---
## 1988 elections analysis

Looks like we dropped some NAs.
```{r}
c(sum(complete.cases(polls_subset)),sum(!complete.cases(polls_subset)))
```

Not ideal; we'll learn about methods for dealing with missing data soon.

Interpretation of results:
  + For a fixed state (or across all states), a non-black male respondent has odds of $e^{0.45}=1.57$ of supporting Bush.

  + For a fixed state and sex, a black respondent as $e^{-1.74}=0.18$ times (an 82% decrease)  the odds of supporting Bush as a non-black respondent; you are much less likely to support Bush if your race is black compared to being non-black.
  
  + For a given state and race, a female respondent has $e^{-0.10}=0.91$ (a 9% decrease)  times the odds of supporting Bush as a male respondent. However, this effect is not actually statistically significant!



---
## 1988 elections analysis

The state-level standard deviation is estimated at 0.41, so that the states do vary some, but not so much.

I expect that you will be able to interpret the corresponding confidence intervals.
```{r echo=F}
confint(model1)
```


---
## 1988 elections analysis

We can definitely fit a more sophisticated model that includes other relevant survey factors, such as

  + region
  
  + prior vote history (note that this is a state-level predictor),
  
  + age, education, and the interaction between them.

Given the structure of the data, it makes sense to include region as a second grouping variable.

We will return to this soon.



---
## 1988 elections analysis

For now, let's just fit two models, one with the main effects for age and education, and the second with the interaction between them.

.small[
```{r, cache=T, echo=F}
model2 <- glmer(bush ~ black + female + edu_label + age_label +(1|state_label),
                family=binomial(link="logit"),data=polls_subset)
print(summary(model2),correlation = FALSE)
```
]

Can you interpret the results?



---
## 1988 elections analysis

```{r, cache=T, results = 'hide'}
model3 <- glmer(bush ~ black + female + edu_label*age_label + (1|state_label),
                family=binomial(link="logit"),data=polls_subset)
```

Looks like we have a convergence issue. These convergence issues are really common when using `glmer`.

Here we have so many parameters to estimate from the interaction terms `edu_label*age_label` (16 actually), and it looks like that's causing a problem.

Now, there are a few potential reasons and fixes for this problem (see [this link](https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html)) but we'll see how we can actually take advantage of the properties of our hierarchical model to get around the issue.

**Side note:** if you suspect your design matrix is not full rank, you can do a quick check using the `rankMatrix` function in the `Matrix` package.



---
## Quick note on estimation

ML estimation is carried out typically using adaptive Gaussian quadrature.

To improve accuracy, many packages (default is usually Laplace approximation) increase the number of quadrature points to be greater than one.

Note that some software packages (including the .hlight[glmer] function in the .hlight[lme4] package) require Laplace approximation with Gaussian quadrature if the number of random effects is more than 1 for the sake of computational efficiency.

The main point though is that it is possible to tweak the approximation, and specifically the optimizer, in the .hlight[glmer] function, so that the usual go-to solution for getting around convergence issues is to simply change the optimizer.

Read more about the [BOBYQA](https://en.wikipedia.org/wiki/BOBYQA) optimizer in particular at your leisure.

**My take:** as I have mentioned before, hierarchical modeling is one of the areas where leaning Bayesian is a huge plus; not having to deal with convergence issues is one of them.




---
## 1988 elections analysis

First, let's go back to the model without the interaction but then try to control for

+ region (since states are nested within regions)
  
+ prior vote history (our state-level predictor),

We have
```{r results = 'hide',cache=T}
model2 <- glmer(bush ~ black + female + v_prev + edu_label + age_label +
                (1|state_label) + (1|region_label),
                family=binomial(link="logit"),data=polls_subset)
```

which also does not converge.


---
## 1988 elections analysis

We are unable to include education and age in this version of the model. Could be that we have too little $\textrm{bush}_i = 1$ or $0$ values for certain combinations? You should check!

As mentioned before, we can actually take advantage of the properties of our hierarchical model to get around the issue.

How about we treat those as varying/random effects instead? Let's try
```{r results = 'hide',cache=T}
model3 <- glmer(bush ~ black + female + v_prev + 
                  (1|state_label) + (1|region_label) + 
                  (1|edu_label:age_label),
                family=binomial(link="logit"),data=polls_subset)
```

This runs fine. Here we are able to borrow information for the combinations of those variables with insufficient data, and that helps a ton!

This is more of an adhoc fix, but it often works really well in practice.

**Side note:** ideally, we should be much more careful with building the model (for example, do we really need to include region?).


---
## 1988 elections analysis

.small[
```{r}
summary(model3)
```
]



---
## 1988 elections analysis

Remember that in the first model, the state-level standard deviation was estimated as 0.41. Looks like we are now able to separate that (for the most part) into state and region effects.

Interpretation of results:

+ For a fixed state, education and age bracket, a non-black male respondent with zero prior average Republican vote share, has odds of $e^{-3.51}=0.03$ of supporting Bush (no one really has 0 value for `v_prev`).
  
+ For a fixed state, sex, education level, age bracket and zero prior average Republican vote share, a black respondent has $e^{-1.75}=0.17$ times (an 83% decrease) the odds of supporting Bush as a non-black respondent, which is about the same as before.
  
+ For each percentage point increase in prior average Republican vote share, residents of a given state, race, sex, education level  age bracket have $e^{0.07}=1.07$ times the odds of supporting Bush.





---

class: center, middle

# What's next?

### Move on to the readings for the next module!




