---
title: "STA 610L: Module 4.5"
subtitle: "Introduction to finite mixture models (categorical data)"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
library(Compositional)
library(MCMCpack)
library(gtools)
```




## Categorical data (univariate)

- Suppose

  + $Y \in \{1,\ldots, D\}$;
  
  + $\Pr(y = d) = \theta_d$ for each $d = 1,\ldots, D$; and
  
  + $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_D)$.
  
- Then the pmf of $Y$ is
.block[
.small[
$$\Pr[y = d| \boldsymbol{\theta}] = \prod_{d=1}^D \theta_d^{\mathbb{1}[y = d]}.$$
]
]

--

- We say $Y$ has a .hlight[multinomial distribution] with sample size 1, or a .hlight[categorical distribution].

--

- Write as $Y | \boldsymbol{\theta} \sim \textrm{Multinomial}(1,\boldsymbol{\theta})$ or $Y | \boldsymbol{\theta} \sim \textrm{Categorical}(\boldsymbol{\theta})$.

--

- Clearly, this is just an extension of the Bernoulli distribution.




---
## Dirichlet distribution

- Since the elements of the probability vector $\boldsymbol{\theta}$ must always sum to one, that is, its support is the $D-1$ .hlight[simplex].

--

- A conjugate prior for categorical/multinomial data is the .hlight[Dirichlet distribution].

--

- A random variable $\boldsymbol{\theta}$ has a .hlight[Dirichlet distribution] with parameter $\boldsymbol{\alpha}$, if
.block[
.small[
$$p[\boldsymbol{\theta} | \boldsymbol{\alpha}] = \dfrac{\Gamma\left(\sum_{d=1}^D \alpha_d\right)}{\prod_{d=1}^D \Gamma(\alpha_d)} \prod_{d=1}^D \theta_d^{\alpha_d-1}, \ \ \ \alpha_d > 0 \ \textrm{ for all } \ d = 1,\ldots, D.$$
]
]

  where $\boldsymbol{\alpha} = (\alpha_1,\ldots,\alpha_D)$, and
.block[
.small[
$$\sum_{d=1}^D \theta_d = 1, \ \ \theta_d \geq 0 \ \textrm{ for all } \ d = 1,\ldots, D.$$
]
]

--

- We write this as $\boldsymbol{\theta} \sim \textrm{Dirichlet}(\boldsymbol{\alpha}) = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_D)$.

--

- The Dirichlet distribution is a multivariate generalization of the .hlight[beta distribution].



---
## Dirichlet distribution

- Write
.block[
.small[
$$\alpha_0 = \sum_{d=1}^D \alpha_d \ \ \ \textrm{and} \ \ \ \alpha_d^\star = \dfrac{\alpha_d}{\alpha_0}.$$
]
]

--

- Then we can re-write the pdf as
.block[
.small[
$$p[\boldsymbol{\theta} | \boldsymbol{\alpha}] = \frac{\Gamma\left(\alpha_0\right)}{\prod_{d=1}^D \Gamma(\alpha_d)} \prod_{d=1}^D \theta_d^{\alpha_d-1}, \ \ \ \alpha_d > 0 \ \textrm{ for all } \ d = 1,\ldots, D.$$
]
]
  
--

- Properties:

  - 
.block[
.small[
$$\mathbb{E}[\theta_d] = \alpha_d^\star;$$
]
]

--

  - 
.block[
.small[
$$\textrm{Mode}[\theta_d] = \dfrac{\alpha_d - 1}{\alpha_0 - d};$$
]
]

--
  
  - 
.block[
.small[
$$\mathbb{Var}[\theta_d] = \dfrac{\alpha_d^\star(1-\alpha_d^\star)}{\alpha_0 + 1} = \dfrac{\mathbb{E}[\theta_d](1-\mathbb{E}[\theta_d])}{\alpha_0 + 1};$$
]
]
  
--

  - 
.block[
.small[
$$\mathbb{Cov}[\theta_d,\theta_k] = \dfrac{\alpha_d^\star\alpha_k^\star}{\alpha_0 + 1} = \dfrac{\mathbb{E}[\theta_d]\mathbb{E}[\theta_k]}{\alpha_0 + 1}.$$
]
]





---
## Dirichlet examples

$\textrm{Dirichlet}(1,1,1)$

```{r fig.height=4.4, echo=F}
draws <- rdirichlet(1000, c(1,1,1))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(10,10,10)$

```{r fig.height=4.4, echo=F}
draws <- rdirichlet(1000, c(10,10,10))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(100,100,100)$

```{r fig.height=4.4, echo=F}
draws <- rdirichlet(1000, c(100,100,100))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(1,10,1)$

```{r fig.height=4.4, echo=F}
draws <- rdirichlet(1000, c(1,10,1))
bivt.contour(draws)
```




---
## Dirichlet examples

$\textrm{Dirichlet}(50,100,10)$

```{r fig.height=4.4, echo=F}
draws <- rdirichlet(1000, c(50,100,10))
bivt.contour(draws)
```





---
## Likelihood

- Let $Y_i, \ldots, Y_n | \boldsymbol{\theta} \sim \textrm{Categorical}(\boldsymbol{\theta})$.

--

- Recall
.block[
.small[
$$\Pr[y_i = d| \boldsymbol{\theta}] = \prod_{d=1}^D \theta_d^{\mathbb{1}[y_i = d]}.$$
]
]

--

- Then,
.block[
.small[
$$
\begin{split}
p[Y | \boldsymbol{\theta}] = p[y_1,\ldots,y_n | \boldsymbol{\theta}] = \prod_{i=1}^n \prod_{d=1}^D \theta_d^{\mathbb{1}[y_i = d]} = \prod_{d=1}^D  \theta_d^{\sum_{i=1}^n \mathbb{1}[y_i = d]} = \prod_{d=1}^D  \theta_d^{n_d}
\end{split}
$$
]
]

  where $n_d$ is just the number of individuals in category $d$.

--

- Maximum likelihood estimate of $\theta_d$ is
.block[
.small[
$$\hat{\theta}_d = \dfrac{n_d}{n}, \ \ d = 1,\ldots, D$$
]
]



---
## Posterior

- Set $\pi(\boldsymbol{\theta}) = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_D)$. Then
.block[
.small[
$$
\begin{split}
\pi(\boldsymbol{\theta} | Y) & \propto p[Y| \boldsymbol{\theta}] \cdot \pi[\boldsymbol{\theta}]\\
& \propto \prod_{d=1}^D  \theta_d^{n_d} \prod_{d=1}^D \theta_d^{\alpha_d-1} \\
& \propto \prod_{d=1}^D  \theta_d^{\alpha_d + n_d - 1}\\
& = \textrm{Dirichlet}(\alpha_1 + n_1,\ldots,\alpha_D + n_D)
\end{split}
$$
]
]

--

- Posterior expectation:
.block[
.small[
$$\mathbb{E}[\theta_d | Y] = \dfrac{\alpha_d + n_d}{\sum_{d^\star=1}^D (\alpha_{d^\star} + n_{d^\star})}.$$
]
]

--

- <div class="question">
We can also extend the Dirichlet-multinomial model to more variables (contingency tables).
</div>

--

- First, what if our data actually comes from $K$ different sub-populations of groups of people?



---
## Finite mixture of multinomials


- For example, if our categorical data comes from men and women, and we don't expect marginal independence across the two groups, then we have a mixture of distributions.

--

- With our data coming from a "combination" or "mixture" of sub-populations, we no longer have independence across all observations, so that the likelihood $p[Y| \boldsymbol{\theta}] \neq \prod\limits_{i=1}^n \prod\limits_{d=1}^D \theta_j^{\mathbb{1}[y_i = d]}$.

--

- However, we can still have "conditional independence" within each group.

--

- Unfortunately, we do not always know the indexes for those groups.

--

- That is, we know our data contains $K$ different groups, but we actually do not know which observations belong to which groups.

--

- **Solution**: introduce a .hlight[latent variable] $z_i$ representing the group/cluster indicator for each observation $i$, so that each $z_i \in \{1,\ldots, K\}$.




---
## Finite mixture of multinomials

- Given the cluster indicator $z_i$ for observation $i$, write
  + $\Pr(y_i = d | z_i) = \psi_{z_i,d} \equiv \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]}$, and
  + $\Pr(z_i = k) = \lambda_k \equiv \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]}$.
  
--

- Then, the marginal probabilities we care about will be
.block[
.small[
$$
\begin{split}
\theta_d & = \Pr(y_i = d)\\
& = \sum_{k=1}^K \Pr(y_i = d | z_i = k) \cdot \Pr(z_i = k)\\
& = \sum_{k=1}^K \lambda_k \cdot \psi_{k,d}, \\
\end{split}
$$
]
]

  which is a .hlight[finite mixture of multinomials], with the weights given by $\lambda_k$.
  
<!-- -- -->

<!-- - Intuitively, if we randomly select any $y_i$, the probability that the $y_i$ is equal to $d$, $\Pr(y_i = d)$, is a weighted average of the probability of $y_i$ is equal to $d$ within each cluster/group. -->




---
## Posterior inference

- Write

  + $\boldsymbol{\lambda} = (\lambda_1, \ldots, \lambda_K)$, and
  
--

  + $\boldsymbol{\psi} = \{\psi_{z_i,d}\}$ to be a $K \times D$ matrix of probabilities, where each $k$th row is the vector of probabilities for cluster $k$.

--

- The observed data likelihood is
.block[
.small[
$$
\begin{split}
p\left[Y = (y_1, \ldots, y_n) | Z = (z_1, \ldots, z_n), \boldsymbol{\psi}, \boldsymbol{\lambda}\right] & = \prod_{i=1}^n \prod\limits_{d=1}^D \Pr\left(y_i = d | z_i, \psi_{z_i,d} \right)\\
& = \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]},
\end{split}
$$
]
]

  which includes products (and not the sums in the mixture pdf), and as you will see, makes sampling a bit easier.
  
--

- Next we need priors.



---
## Posterior inference

- First, for $\boldsymbol{\lambda} = (\lambda_1, \ldots, \lambda_K)$, the vector of cluster probabilities, we can use a Dirichlet prior. That is,
.block[
.small[
$$\pi[\boldsymbol{\lambda}] = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_K) \propto \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1}.$$
]
]
  
--

- For $\boldsymbol{\psi}$, we can assume independent Dirichlet priors for each cluster vector $\boldsymbol{\psi}_k = (\psi_{k,1}, \ldots, \psi_{k,D})$. That is, for each $k = 1, \ldots, K$, 
.block[
.small[
$$\pi[\boldsymbol{\psi}_k] = \textrm{Dirichlet}(a_1,\ldots,a_d) \propto \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1}.$$
]
]
  
--

- Finally, from our distribution on the $z_i$'s, we have
.block[
.small[
$$
\begin{split}
p\left[Z = (z_1, \ldots, z_n) | \boldsymbol{\lambda}\right] & = \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]}.
\end{split}
$$
]
]




---
## Posterior inference

- Note that the unobserved variables and parameters are $Z = (z_1, \ldots, z_n)$, $\boldsymbol{\psi}$, and $\boldsymbol{\lambda}$.

--

- So, the joint posterior is
.block[
.small[
$$
\begin{split}
\pi\left(Z, \boldsymbol{\psi}, \boldsymbol{\lambda} | Y \right) & \propto p\left[Y | Z, \boldsymbol{\psi}, \boldsymbol{\lambda}\right]  \cdot p(Z| \boldsymbol{\psi}, \boldsymbol{\lambda})  \cdot \pi(\boldsymbol{\psi}, \boldsymbol{\lambda}) \\
\\
& \propto \left[ \prod_{i=1}^n \prod\limits_{d=1}^D p\left(y_i = d | z_i, \psi_{z_i,d} \right) \right]  \cdot p(Z| \boldsymbol{\lambda})  \cdot \pi(\boldsymbol{\psi}) \cdot \pi(\boldsymbol{\lambda}) \\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]} \right) \\
& \ \ \ \ \ \times \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \\
& \ \ \ \ \ \times \left( \prod\limits_{k=1}^K \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \\
& \ \ \ \ \ \times \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right). \\
\end{split}
$$
]
]



---
## Posterior inference

- First, we need to sample the $z_i$'s, one at a time, from their full conditionals. 

--

- For $i = 1, \ldots, n$, sample $z_i \in \{1,\ldots, K\}$ from a categorical distribution (multinomial distribution with sample size one) with probabilities
.block[
.small[
$$
\begin{split}
\Pr[z_i = k | \dots ] & = \Pr[z_i = k | y_i, \boldsymbol{\psi}_k, \lambda_k] \\
\\
& = \dfrac{ \Pr[y_i, z_i = k | \boldsymbol{\psi}_k, \lambda_k] }{ \sum\limits^K_{l=1} \Pr[y_i, z_i = l | \boldsymbol{\psi}_l, \lambda_l] } \\
\\
& = \dfrac{ \Pr[y_i | z_i = k, \boldsymbol{\psi}_k] \cdot \Pr[z_i = k, \lambda_k] }{ \sum\limits^K_{l=1} \Pr[y_i | z_i = l, \boldsymbol{\psi}_l] \cdot \Pr[z_i = l, \lambda_l] } \\
\\
& = \dfrac{ \psi_{k,d} \cdot \lambda_k }{ \sum\limits^K_{l=1} \psi_{l,d} \cdot \lambda_l }. \\
\end{split}
$$
]
]



---
## Posterior inference

- Next, sample each cluster vector $\boldsymbol{\psi}_k = (\psi_{k,1}, \ldots, \psi_{k,D})$ from
.block[
.small[
$$
\begin{split}
\pi[\boldsymbol{\psi}_k | \dots ] & \propto \pi\left(Z, \boldsymbol{\psi}, \boldsymbol{\lambda} | Y \right) \\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]} \right) \cdot \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \cdot \left( \prod\limits_{k=1}^K \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left(  \prod\limits_{d=1}^D \psi_{k,d}^{n_{k,d}} \right) \cdot \left( \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \\
\\
& = \left(  \prod\limits_{d=1}^D \psi_{k,d}^{a_d + n_{k,d} - 1} \right) \\
\\
& \equiv \textrm{Dirichlet}\left(a_1 + n_{k,1},\ldots,a_D + n_{k,D}\right).
\end{split}
$$
]
]

  where $n_{k,d} = \sum\limits_{i: z_i = k} \mathbb{1}[y_i = d]$, the number of individuals in cluster $k$ that are assigned to category $d$ of the levels of $y$.



---
## Posterior inference

- Finally, sample $\boldsymbol{\lambda} = (\lambda_1, \ldots, \lambda_K)$, the vector of cluster probabilities from
.block[
.small[
$$
\begin{split}
\pi[\boldsymbol{\lambda} | \dots ] & \propto \pi\left(Z, \boldsymbol{\psi}, \boldsymbol{\lambda} | Y \right) \\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{d=1}^D \psi_{z_i,d}^{\mathbb{1}[y_i = d | z_i]} \right) \cdot \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \cdot \left( \prod\limits_{k=1}^K \prod\limits_{d=1}^D \psi_{k,d}^{a_d-1} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left( \prod_{i=1}^n \prod\limits_{k=1}^K \lambda_k^{\mathbb{1}[z_i = k]} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left( \prod\limits_{k=1}^K \lambda_k^{n_k} \right) \cdot \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k-1} \right)\\
\\
& \propto \left( \prod\limits_{k=1}^K \lambda_k^{\alpha_k + n_k - 1} \right)\\
& \equiv \textrm{Dirichlet}\left(\alpha_1 + n_1,\ldots,\alpha_K + n_K\right),\\
\end{split}
$$
]
]

  with $n_k = \sum\limits_{i=1}^n \mathbb{1}[z_i = k]$, the number of individuals assigned to cluster $k$.



---
## Categorical data: bivariate case

- How can we extend the same ideeas to multiple categorical variables?

--
  
- Well let's start small. Suppose we have data $(y_{i1},y_{i2})$, for $i = 1, \ldots, n$, where
  + $y_{i1} \in \{1,\ldots, D_1\}$
  + $y_{i2} \in \{1,\ldots, D_2\}$.
  
--

- This is just a two-way contingency table, so that we are interested in estimating the probabilities $\Pr(y_{i1} = d_1, y_{i2} = d_2) = \theta_{d_1d_2}$.
    
--
      
- Write $\boldsymbol{\theta} = \{\theta_{d_1d_2}\}$, which is a $D_1 \times D_2$ matrix of all the probabilities.



---
## Categorical data: bivariate case
        
- The likelihood is therefore
.block[
.small[
$$
\begin{split}
p[Y| \boldsymbol{\theta}] & = \prod_{i=1}^n \prod_{d_2=1}^{D_2} \prod_{d_1=1}^{D_1} \theta_{d_1d_2}^{\mathbb{1}[y_{i1} = d_1, y_{i2} = d_2]}\\
\\
& = \prod_{d_2=1}^{D_2} \prod_{d_1=1}^{D_1} \theta_{d_1d_2}^{\sum\limits_{i=1}^n \mathbb{1}[y_{i1} = d_1, y_{i2} = d_2]}\\
\\
& = \prod_{d_2=1}^{D_2} \prod_{d_1=1}^{D_1} \theta_{d_1d_2}^{n_{d_1d_2}}
\end{split}
$$
]
]

  where $n_{d_1d_2} = \sum\limits_{i=1}^n \mathbb{1}[y_{i1} = d_1, y_{i2} = d_2]$ is just the number of observations in cell $(d_1,d_2)$ of the contingency table.
      
      
      
---
## Posterior inference
        
- How can we do Bayesian inference?

--

- Several options! Most common are:

--
        
- .hlight[Option 1:] Follow the univariate approach.

  + Rewrite the bivariate data as univariate data, that is, $y_i \in \{1,\ldots, D_1 D_2\}$.

--

  + Write $\Pr(y_i = d) = \nu_d$ for each $d = 1,\ldots, D_1 D_2$.
  
--

  + Specify Dirichlet prior as $\boldsymbol{\nu} = (\nu_1,\ldots,\nu_{D_1 D_2}) \sim \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_{D_1 D_2})$.
  
--

  + Then, posterior is also Dirichlet with parameters updated with the number in each cell of the contingency table.
      


---
## Posterior inference

- .hlight[Option 2:] Assume independence, then follow the univariate approach.

  + Write $\Pr(y_{i1} = d_1, y_{i2} = d_2) = \Pr(y_{i1} = d_1)\Pr(y_{i2} = d_2)$, so that $\theta_{d_1d_2} = \lambda_{d_1} \psi_{d_2}$.
  
--

  + Specify independent Dirichlet priors on $\boldsymbol{\lambda} = (\lambda_1,\ldots,\lambda_{D_1})$ and $\boldsymbol{\psi} = (\psi_1,\ldots,\psi_{D_2})$.
  
--

  + That is,
      + $\boldsymbol{\lambda} \sim \textrm{Dirichlet}(a_1,\ldots,a_{D_1})$
      
      + $\boldsymbol{\psi} \sim \textrm{Dirichlet}(b_1,\ldots,b_{D_2})$.

--
  + This reduces the number of parameters from $D_1 D_2 - 1$ to $D_1 + D_2 - 2$.
      

---
## Posterior inference
        
- .hlight[Option 3:] Log-linear model
  + $\theta_{d_1d_2} = \dfrac{e^{ \alpha_{d_1} + \beta_{d_2} + \gamma_{d_1d_2} }}{ \sum\limits_{d_2=1}^{D_2} \sum\limits_{d_1=1}^{D_1} e^{ \alpha_{d_1} + \beta_{d_2} + \gamma_{d_1d_2} }}$;
  
--

  + Specify priors (perhaps normal) on the parameters.
      


---
## Posterior inference
        
- .hlight[Option 4:] Latent structure model
  + Assume conditional independence given a .hlight[latent variable];
  
--

  + That is, write
  .block[
.small[
$$
\begin{split}
\theta_{d_1d_2} & = \Pr(y_{i1} = d_1, y_{i2} = d_2)\\
& = \sum_{k=1}^K \Pr(y_{i1} = d_1, y_{i2} = d_2 | z_i = k) \cdot \Pr(z_i = k)\\
& = \sum_{k=1}^K \Pr(y_{i1} = d_2| z_i = k) \cdot \Pr(y_{i2} = d_2 | z_i = k) \cdot \Pr(z_i = k)\\
& = \sum_{k=1}^K \lambda_{k,d_1} \psi_{k,d_2} \cdot \omega_k .\\
\end{split}
$$
]
]

--

  + This is once again, a .hlight[finite mixture of multinomial distributions].
     


---
## Categorical data: extensions

- For categorical data with more than two categorical variables, it is relatively easy to extend the framework for latent structure models.

--

- Clearly, there will be many more parameters (vectors and matrices) to keep track of, depending on the number of clusters and number of variables!

--

- If interested, read up on .hlight[finite mixture of products of multinomials]. 

--

- Can also go full Bayesian nonparametrics with a .hlight[Dirichlet process mixture of products of multinomials]. 

--

- Happy to provide resources for those interested!





---

class: center, middle

# What's next?

### Move on to the readings for the next module!




